A Practical Optional Type System for Clojure

### 2012-10-24 dissertation/design-choices.tex
Add proposal, acknowledgements

Strickland, Tobin-Hochstadt, and Felleisen's describe variable-arity polymorphism~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al./ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).

@@

\section{Clojure Hierarchy}

The Clojure hierarchy is a set of Java classes and interfaces, mostly in the
\emph{clojure.lang} package, that define Clojure's core 
interfaces and data structures.
Typed Clojure parameterises some of these classes in a way that resembles
Java or Scala generics.

Clojure's hierarchy was accommodating to this process.
For example, \lstinline{clojure.lang.Seqable} is the main interface behind
Clojure's sequence abstraction. Sequences are immutable collections,
so it was natural to parameterise \lstinline{Seqable} with one covariant
argument (covariant means \lstinline|(Seqable Integer)| is
a subtype of \lstinline|(Seqable Number)| because \lstinline|Integer|
is a subtype of \lstinline|Number|).
All ``seqable'' classes in the Clojure hierarchy implement \lstinline|Seqable|,
so updating them to use the parameterised \lstinline|Seqable| was a matter of
searching for classes that had \lstinline|Seqable| as an ancestor and replacing
the unparameterised \lstinline|Seqable| with the corresponding parameterised 
version.

@@

Type inference in Typed Clojure is based on Local Type Inference
by Pierce and Turner~\cite{PT00}.
The main extension to Local Type Inference in Typed Clojure
is support for \emph{dotted variables} for variable-arity polymorphism.
This implementation is lifted directly from Typed Racket.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys.
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

Heterogeneous map types do not support concatenation (merging), as they
only hold positive information on the keys with entries, not
which keys are \emph{absent}.
This makes concatenation unsafe, as keys to the right of a
concatenation can ``overwrite'' keys to the left. With no information
on the absence of keys, concatenation becomes an unsound operation.

>>>>

Strickland, Tobin-Hochstadt, and Felleisen describe variable-arity poly\-morphism ~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al.\ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).

@@

\section{Clojure type hierarchy}

The Clojure type hierarchy is a set of Java classes and interfaces, mostly in the
\emph{clojure.lang} package, that define Clojure's core 
interfaces and data structures.
Typed Clojure parameterises some of these classes in a way that resembles
Java or Scala generics.

Clojure's type hierarchy was accommodating to this process.
For example, \lstinline{clojure.lang.Seqable} is the main interface behind
Clojure's sequence abstraction. Sequences are immutable collections,
so it was natural to parameterise \lstinline{Seqable} with one covariant
argument (covariant means \lstinline|(Seqable Integer)| is
a subtype of \lstinline|(Seqable Number)| because \lstinline|Integer|
is a subtype of \lstinline|Number|).
All ``seqable'' classes in the Clojure hierarchy implement \lstinline|Seqable|,
so updating them to use the parameterised \lstinline|Seqable| was a matter of
searching for classes that had \lstinline|Seqable| as an ancestor and replacing
the unparameterised \lstinline|Seqable| with the corresponding parameterised 
version.

@@

Type inference in Typed Clojure is based on Local Type Inference
by Pierce and Turner~\cite{PT00}.
The main extension to Local Type Inference in Typed Clojure
is support for \emph{dotted variables} for variable-arity polymorphism.
This implementation is ported directly from Typed Racket.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys with the type associated with each key in
a map tracked separately, similar to type systems for record types 
(see work by Harper and Pierce~\cite{HP91}).
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

Heterogeneous map types do not support concatenation (merging), as they
only hold positive information on the keys with entries, not
which keys are \emph{absent}.
This makes concatenation unsafe, as keys to the right of a
concatenation can ``overwrite'' keys to the left with different types. With no information
on the absence of keys, concatenation becomes an unsound operation.

### 2012-10-24 dissertation/experiments.tex
Add proposal, acknowledgements

In this chapter we look at several example of using Typed Clojure and
how well the current prototype handles them. 
We intentially chose examples that could be challenging to type check
or were particularly useful to the everyday Clojure programmer.

@@

The initial motivation for adding user defined type constructors to Typed Clojure
(as discussed in section \ref{sec:monaddef}) was to type check
monad transformer definitions.
Monad transformers are always parameterised by a monad type constructor, and should work for all
monad type constructors. Keeping the monad type constructor abstract allows us to determine
whether a monad transformer works \emph{for any} monads.

@@

Conduits are an advanced form of ``pipes'' using arrows, and a generalisation of monads~\cite{ConduitRef}.
\emph{Conduit} is a Clojure library developed by Duey and 
contributors~\footnote{Github home of Conduit: \url{https://github.com/jduey/conduit}} 
that supports programming with conduits.

@@

An interesting property of the types assigned to conduits (which
are internally functions of a single argument) is that it is
a recursive type. This is not in itself interesting, but Typed Clojure
implements conduit types as \emph{type functions} with variance,
which required the variance of the conduit types to be known in advance.

@@

The resulting port of \emph{Conduit} was unsatisfying for real world use.
The simplifications made in order to type check the library reversed
any effort \emph{Conduit} made to conform to common Clojure idioms.
The future work on improving variable-argument types should reveal whether
this library can be ported satisfactorily.

>>>>

In this chapter work through several examples of using Typed Clojure and
how well the current prototype handles them. 
We intentially chose examples that could be challenging to type check
or were particularly useful to the everyday Clojure programmer.

@@

The initial motivation for adding user defined type constructors to Typed Clojure
(as discussed in section \ref{sec:monaddef}) was to type check
monad transformer definitions.
Monad transformers are always parameterised by a monad type constructor, and should work for all
monad type constructors. Keeping the monad type constructor abstract allows us to determine
whether a monad transformer works \emph{for all} monads.

@@

Conduits are an advanced form of ``pipes'' using arrows, a generalisation of monads~\cite{ConduitRef}.
\emph{Conduit} is a Clojure library developed by Duey and 
contributors~\footnote{Github home of Conduit: \url{https://github.com/jduey/conduit}} 
that supports programming with conduits.

@@

An interesting property of the types assigned to conduits (which
are internally functions of a single argument) is that they are 
recursive types. This is not in itself interesting, but Typed Clojure
implements conduit types as \emph{type functions} with variance,
which required the variance of the conduit types to be known in advance.

@@

The resulting port of \emph{Conduit} was unsatisfying for real world use.
The simplifications made in order to type check the library reversed
any effort that was invested into \emph{Conduit} to conform to common Clojure idioms.
The future work on improving variable-argument types should reveal whether
this library can be ported satisfactorily. It is desirable to port
this library because implementations and usages of conduits result in very abstract code
which static type systems can help verify as correct.

### 2012-10-24 dissertation/future-work.tex
final dissertation (unreviewed by Rowan)
Add proposal, acknowledgements

\section{Paired Arguments}

@@

A key part of Typed Racket~\cite{Tob10} is its contract and blame system.
They enable safe interoperability with untyped Racket by generating runtime
contracts in key places based on static types. Typed Racket also 
includes a sophisticated blame system based on Wadler\cite{WF09}
to provide more accurate error reporting that ensures typed modules
``can't be blamed''.

@@

It would be desirable to port Typed Clojure to other Clojure implementations.
Each implementation would bring its own set of challenges with interoperability.
The core of Typed Clojure would preserved with each port changing at least 
its interoperability with its host platform. Section \label{sec:portability}
discusses potential problems with the porting process, and how
they might be solved.

>>>>

\section{Variable-Arity Polymorphism}
\label{sec:futurearity}

@@

A key part of Typed Racket~\cite{Tob10} is its contract and blame systems.
They enable safe interoperability with untyped Racket by generating runtime
contracts in key places based on static types. Typed Racket also 
includes a sophisticated blame system based on Wadler\cite{WF09}
to provide more accurate error reporting that ensures typed modules
``can't be blamed''.

@@

It would be desirable to port Typed Clojure to other Clojure implementations.
Each implementation would bring its own set of challenges with interoperability.
The core of Typed Clojure would preserved with each port changing at least 
its interoperability with its host platform. Section \ref{sec:portability}
discusses potential problems with the porting process, and how
they might be solved.

### 2012-10-24 dissertation/implementation.tex
Hmap design
final dissertation (unreviewed by Rowan)
Add proposal, acknowledgements

This chapter discusses the implementation of the Typed Clojure prototype type 
system (we refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter)
in some detail concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follow the implementation of Typed Racket,
which is reflected in this chapter.

@@

\lstinline|infer| is almost always used when invoking polymorphic functions.
\lstinline|clojure.core/constantly| is a polymorphic function with type \lstinline|(All [x y] [x -> [y * -> x]])|
(read as a function taking \lstinline|x| and returning a function that takes any number
of \lstinline|y|'s and returns \lstinline|x|).
For instance, \lstinline|((constantly true) 'any 'number)| results in the value \lstinline|true|.

@@

\section{Variable-arity Polymorphism}

@@

\section{Portability to other Clojure Dialects}

>>>>

This chapter discusses the implementation of the Typed Clojure prototype type 
system (we refer to this implementation as \emph{Typed Clojure} for the remainder of this chapter)
in some detail concentrating on significant challenges that were identified and overcome.
Many aspects of Typed Clojure's design follow the implementation of Typed Racket,
which is reflected in this chapter. 
(The code is available on Github at \url{https://github.com/frenchy64/typed-clojure}).

@@

\lstinline|infer| is almost always used when invoking polymorphic functions like,
for example, \lstinline|constantly|, which has type \lstinline|(All [x y] [x -> [y * -> x]])|
(read as a function taking \lstinline|x| and returning a function that takes any number
of \lstinline|y|'s and returns \lstinline|x|).
For instance, \lstinline|((constantly true) 'any 'number)|, results in the value \lstinline|true|.

@@

\section{Variable-arity Polymorphism}
\label{sec:impvariablearity}

@@

\section{Portability to other Clojure Dialects}
\label{sec:portability}

### 2012-10-24 dissertation/paper.tex
Half of Rowan's corrections
Add proposal, acknowledgements

\keywords{Type systems, Clojure}
% TODO What are categories? cshonours requires them
\categories{}

@@

@@

Typed Clojure is intended to be of practical use to Clojure programmers.
We show that Typed Clojure help verify Clojure code written in sophisticated
programming styles is correct by porting most of a wide-spread Clojure library for monadic programming.
Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
a useful tool when code involves usages of Java's \emph{null}.

>>>>

\keywords{Type systems, Clojure}
% TODO What are categories? cshonours requires them - but commented out now
%\categories{}

@@

\begin{acknowledgements}

The past year has included by far my most enjoyable and motivating programming-related experiences.

Last November I attended my first conference, Clojure Conj 2011 (Raleigh, NC),
where I delivered a talk on logic programming with Clojure and met some of my programming heroes.
A week later my supervisor Rowan Davies invited me to visit Carnegie-Mellon University
where he was working on his PhD project. Most memorably, I was invited to a meeting between
Rowan and his PhD supervisor Frank Pfenning, which was a great thrill.

This trip was funded by donations from the programming community. Thank you to all who helped me get there:
this work was directly inspired from conversations during this period.

Many thanks go to my supervisor Rowan. His influence as a lecturer inspired my exploration of
functional programming languages. He has exceeded my expectations as an honours supervisor,
often dedicating time to assist me on public holidays and weekends.
Rowan's experience with refinement types brought a unique perspective to this project
which was very welcome.

I'd also like to thank: Sam Tobin-Hochstadt for his consistent assistance and patience;
David Nolen for mentoring this project as part of Google Summer of Code 2012;
and Chas Emerick, without whom I would have never considered attending Clojure Conj 2011
(and this work probably would not have happened).
Also, Daniel Spiewak, Dan Friedman and William Byrd, Jim Duey, Baishampayan Ghose, and
Steve Strickland, thank you for your help and inspiration.
Thank you Rich Hickey for creating Clojure, it is a thrill to build tools that make
this great language even better.

Fittingly, I have been invited to talk on Typed Clojure at Clojure Conj
2012 in November, which I am highly anticipating.

Finally, thanks to my family, friends and to my girlfriend Tamara for your love and support.

\end{acknowledgements}

@@

Typed Clojure is intended to be of practical use to Clojure programmers.
We show that Typed Clojure helps verify the absence of errors in Clojure code written in sophisticated
programming styles by porting most of a wide-spread Clojure library for monadic programming.
Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
a useful tool to show misusing of Java's \emph{null}.
